---
title: "EDA"
output: html_notebook
---

Import libraries
```{r warning=FALSE}
library(tidyverse)
```


## Read in Data
```{r}
df <- read.csv("cafo-wq-data.csv")
```

## Initial EDA:
In order to construct an analysis plan and propose an appropriate model, we need to preform some exploratory data analysis.

Data structure

A lot of columns- do we want to/can we filter down to one spatial scale?
```{r}
str(df)
```
Check for missing values
```{r}
sum_na <- function(column){
 sum(is.na(column))
}

df |> summarize(across(everything(), sum_na))
```

Should response variables be log transformed?
Histograms of raw & log transformed continuous response variables
```{r}
# histograms of raw response variables
par(mfrow = c(2, 3))

hist(df$fecalcoliformcol100ml,
     main = 'Fecal coliform',
     xlab = 'Count',
     breaks = 30)

hist(df$no2no3mgl,
     main = 'Nitrate/nitrite',
     xlab = 'mg/L',
     breaks = 30)

hist(df$phosphorusmgl,
     main = 'Phosphorus',
     xlab = 'mg/L',
     breaks = 30)

# histograms of log-transformed response variables
hist(df$feclog,
     main = 'Log fecal coliform',
     xlab = 'log value',
     breaks = 30)

hist(df$nitlog,
     main = 'Log nitrate/nitrite',
     xlab = 'log value',
     breaks = 30)

hist(df$phoslog,
     main = 'Log phosphorus',
     xlab = 'log value',
     breaks = 30)
```

