---
title: "Final Models"
output: html_notebook
---

In this notebook, we will fit the final LMER and GAMM models.

Import libraries
```{r warning=FALSE}
library(lme4)
library(tidyverse)
library(performance)

# for printing and exporting model info
library(modelsummary)
library(gt)

# supress scientific notation in outputs
options(scipen = 999)
```

Read in and clean data
```{r}
setwd("~/Downloads/ST 542/ST542 Project")
raw_data <- read.csv("cafo-wq-data-rev.csv")

# format date column
raw_data$date <- as.Date(raw_data$date, format = '%m/%d/%y')
  
# extract month from date
raw_data <- raw_data |>
  mutate(month = month(date))
  
# add wet/dry season indicator
wet <- c(5, 6, 7, 8, 9)
  
raw_data$szn <- ifelse(raw_data$month %in% wet, 'wet', 'dry')
raw_data$szn <- factor(raw_data$szn, levels = c('dry', 'wet'))
  
# convert columns to factors
raw_data <- raw_data |>
  mutate(
    huc8id = as.factor(huc8id),
    huc10id = as.factor(huc10id),
    huc12id = as.factor(huc12id),
    station_id = as.factor(station_id),
    stationcategory = as.factor(stationcategory),
    month = as.factor(month)
  )

# scale CAFO metric variables
huc12_df <- raw_data |>
  mutate(
    # all CAFO metric
    huc12_nearest_upstream_cafo_dist_scaled = huc12_nearest_upstream_cafo_dist / 1000,
    
    # poultry CAFO metric
    huc12_nearest_p_upstream_cafo_dist_scaled = huc12_nearest_upstream_poultry_c / 1000,
    
    # swine CAFO metrics
    huc12_upstream_s_cafo_density_scaled = huc12_upstream_swine_cafo_densit / 0.1,
    huc12_upstream_s_cafo_count_scaled = huc12_upstream_swine_cafo_count / 10
  )
```

## LMER Models: All CAFO Metrics

Log Nitrite/Nitrate model
```{r}
nitlog_lmer <- lmer(nitlog ~ huc12_nearest_upstream_cafo_dist_scaled
                    + stationcategory
                    + month
                    + (1 | huc8id / station_id),
                    data = huc12_df,
                    REML = TRUE)

summary(nitlog_lmer)
print(paste("AIC:", AIC(nitlog_lmer)))
r2(nitlog_lmer)
```

Log Fecal Coliform Counts
```{r}
feclog_lmer <- lmer(feclog ~ huc12_nearest_upstream_cafo_dist_scaled
                    + huc12_upstream_other_npdes_densi
                    + stationcategory
                    + month
                    + (1 | huc8id/station_id),
                    data = huc12_df,
                    REML = TRUE)

summary(feclog_lmer)
print(paste("AIC:", AIC(feclog_lmer)))
r2(feclog_lmer)
```


Log Phosphorous concentration
```{r}
phoslog_lmer <- lmer(phoslog ~ huc12_nearest_upstream_cafo_dist_scaled
                    + huc12_pop_density
                    + huc12_upstream_wastewater_densit
                    + stationcategory
                    + month
                    + (1 | huc8id/station_id),
                    data = huc12_df,
                    REML = TRUE)

summary(phoslog_lmer)
print(paste("AIC:", AIC(phoslog_lmer)))
r2(phoslog_lmer)
```

Print and export model info in a table
```{r}
tbl <- modelsummary(
  list(
    'Nitrate' = nitlog_lmer,
    'Fecal Coliform' = feclog_lmer,
    'Phosphorus' = phoslog_lmer
  ),
  statistic = '({std.error})',
  stars = TRUE,
  coef_omit = '^(Intercept)|^month',
  output = 'gt'
)

tbl

# save table
gtsave(tbl, 'LMER_all_cafo_results.docx')
```


## LMER Models: Poultry Metrics

Log Nitrite/Nitrate Concentration
```{r}
nitlog_lmer_p <- lmer(nitlog ~ huc12_nearest_p_upstream_cafo_dist_scaled
                      + month
                      + (1 | huc8id/station_id),
                      data = huc12_df,
                      REML = TRUE)

summary(nitlog_lmer_p)
print(paste("AIC:", AIC(nitlog_lmer_p)))
r2(nitlog_lmer_p)
```

Log Fecal Coliform Counts

feclog ~ huc12_nearest_p_upstream_cafo_dist_scaled + huc12_pct_cropland +  
    huc12_weighted_landfill_load + huc12_upstream_other_npdes_densi +      month + (1 | huc8id/station_id)
```{r}
feclog_lmer_p <- lmer(feclog ~ huc12_nearest_p_upstream_cafo_dist_scaled
                      + huc12_pct_cropland
                      + huc12_weighted_landfill_load
                      + huc12_upstream_other_npdes_densi
                      + month
                      + (1 | huc8id/station_id),
                      data = huc12_df,
                      REML = TRUE)

summary(feclog_lmer_p)
print(paste("AIC:", AIC(feclog_lmer_p)))
r2(feclog_lmer_p)
```


Log Phosphorous concentration
```{r}
phoslog_lmer_p <- lmer(phoslog ~ huc12_nearest_p_upstream_cafo_dist_scaled
                       + huc12_upstream_s_cafo_density_scaled
                       + huc12_upstream_s_cafo_count_scaled
                       + huc12_pop_density
                       + huc12_pct_cropland
                       + huc12_upstream_wastewater_densit
                       + stationcategory
                       + month
                       + (1 | huc8id/station_id),
                       data = huc12_df,
                       REML = TRUE)

summary(phoslog_lmer_p)
print(paste("AIC:", AIC(phoslog_lmer_p)))
r2(phoslog_lmer_p)
```

Print and export model info in one table
```{r}
tbl <- modelsummary(
  list(
    'Nitrate' = nitlog_lmer_p,
    'Fecal Coliform' = feclog_lmer_p,
    'Phosphorus' = phoslog_lmer_p
  ),
  statistic = '({std.error})',
  stars = TRUE,
  coef_omit = '^(Intercept)|^month',
  output = 'gt'
)

tbl

# save table
gtsave(tbl, 'LMER_poultry_cafo_results.docx')
```

