---
title: "modeling_Sarat"
author: "Sarat Bantupalli"
date: "2026-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,
                      warning = FALSE, message = FALSE)
```

This is Sarat's progress on modeling the CAFO's data.

# Libraries
We need the following R packages to run this project.

```{r}
library(tidyverse) # for data manipulation
library(lubridate) # for working with datetime objects
```

# Data
Here the data is read into the R environment. Also, let us re-arrange the order of the *date* variable in the data.

```{r}
cafo_data <- read_csv("cafo-wq-data.csv") %>% 
  relocate(date, year, station_id, latitude, longitude, 
           feclog, no2no3mgl, phosphorusmgl, fecpos,
           exceed_fec, exceed_nit, exceed_phos) %>% 
  mutate(date = mdy(date)) %>% 
  rename(fecal_coliform_count = feclog, total_nitrogen_conc = no2no3mgl, 
         total_phosphorus_conc = phosphorusmgl, 
         fecal_coliform_detected = fecpos,
         fecal_coliform_exceeded = exceed_fec,
         total_N_exceeded = exceed_nit, 
         total_P_exceeded = exceed_phos)
cafo_data
```

We can see that we have `r nrow(cafo_data)` observations, and `r ncol(cafo_data)` in our dataset. Let us re-arrange the order of the variables first.

## Potential response variables
The potential response variables of interest include:  

 + fecal_coliform_count: Fecal coliform count, log10 transformed (log10-CFU/100 mL)  
 + total_nitrogen_conc: Nitrate/nitrite concentration (combined as N) (mg/L)  
 + total_phosphorus_conc: Phosphorus concentration (mg/L)  
 + fecal_coliform_detected: a binary variable that represents if Fecal coliform is detected  
 + fecal_coliform_exceeded : a binary variable that represents if Fecal coliform exceeds 200 CFU/100 mL  
 + total_N_exceeded : a binary variable that represents if Nitrate/nitrite (combined as N) exceeds 1 mg/L  
 + total_P_exceeded : a binary variable that represents if Phosphorus exceeds 0.1 mg/L  

## Potential co-variates



