---
title: "Variable Selection"
output: html_notebook
---

Library & data imports

```{r warning=FALSE}
library(lme4)
library(lmerTest)
library(tidyverse)

df <- read.csv("cafo-wq-data.csv")
```

## Data cleaning

```{r}
# format date column
df$date <- as.Date(df$date, format = '%m/%d/%y')

# extract month from date as new column
df <- df |>
  mutate(month = month(date))

# convert columns to factors where necessary
df <- df |>
  mutate(
    huc8id = as.factor(huc8id),
    huc10id = as.factor(huc10id),
    huc12id = as.factor(huc12id),
    station_id = as.factor(station_id),
    stationcategory = as.factor(stationcategory),
    month = as.factor(month)
  )
```

### Helper function to run stepwise selection

This will select the covariates and random effects to include in a model for each predictor. Then, it will display the model comparison metrics for the selected model for each predictor so that we can determine which predictor to use.

Running stepwise selection on all predictors and all covariates at the same time results in only covariates being selected for the final model, despite including a predictor resulting in better model comparison metrics.

This can be reused for model selection across all spatial scales and response variables by changing the function inputs.

cafo_model_selection <- function(data, response, predictors, confounding_vars) {
  
  full_models <- list()
  step_results <- list()
  selected_models <- list()
  
  for (predictor in predictors) {
    
    fixed_effects <- paste(
      c(predictor, confounding_vars),
      collapse = ' + '
    )
    
    model_formula <- as.formula(
      paste(
        response,
        '~',
        fixed_effects,
        '+ (1 | huc8id / station_id)'
      )
    )
    
    # since we are doing model comparison, use REML = FALSE
    # when fitting final models it will be changed to REML = TRUE
    full_model <- lmer(
      model_formula,
      data = data,
      REML = FALSE
    )
    
    # run stepwise selection
    step_result <- step(
      full_model,
      keep = predictor
    )
    
    selected_model <- get_model(step_result)
    
    full_models[[predictor]] <- full_model
    step_results[[predictor]] <- step_result
    selected_models[[predictor]] <- selected_model
  }
  
  # print evaluation metrics for selected model for each predictor
  model_summary <- data.frame(
    response = response,
    predictor = names(selected_models),
    AIC = sapply(selected_models, AIC),
    BIC = sapply(selected_models, BIC),
    logLik = sapply(selected_models, function(x) as.numeric(logLik(x)))
  ) |>
    arrange(AIC)
  
  # store best model
  best_predictor <- model_summary$predictor[1]
  best_model <- selected_models[[best_predictor]]
  
  list(
    response = response,
    full_models = full_models,
    selected_models = selected_models,
    model_summary = model_summary,
    best_predictor = best_predictor,
    best_model = best_model
  )
}
```

## Stepwise selection: All CAFO metrics, spatial scale = huc12

First we will do stepwise selection for all three of the continuous log transformed response variables using the all CAFO types predictor variables.

Scale predictor variables in a way that will make model coefficients more interpretable. Scales were chosen based on range of values for each predictor. Without scaling, models give warnings about variables being on different scales.

Interpretation of coefficients after scaling: expected change in the response for each X unit increase in {predictor variable}

Confounders are currently not being scaled as scaling them had no effect on the models

We also need to remove missing values from predictors that are being tested or else we will get an error when running the stepwise selection

```{r}
huc12_df <- df |>
  drop_na(huc12_nearest_upstream_cafo_dist)

# scale predictors
huc12_df <- huc12_df |>
  mutate(
    huc12_upstream_cafo_density_scaled = huc12_upstream_cafo_density / 0.1,
    huc12_upstream_cafo_count_scaled = huc12_upstream_cafo_count / 10,
    huc12_weighted_cafo_load_scaled = huc12_weighted_cafo_load / 10,
    huc12_nearest_upstream_cafo_dist_scaled = huc12_nearest_upstream_cafo_dist / 1000
  )
```

Define predictors and covariates to test in stepwise selection

```{r}
# scaled cafo predictors to test one at a time
all_cafo_predictors <- c(
  'huc12_upstream_cafo_density_scaled',
  'huc12_upstream_cafo_count_scaled',
  'huc12_weighted_cafo_load_scaled',
  'huc12_nearest_upstream_cafo_dist_scaled'
)

# confounders for backward elimination
confounders <- c(
  'huc12_pop_density',
  'huc12_pct_cropland',
  'huc12_upstream_wastewater_count',
  'huc12_upstream_landfill_count',
  'huc12_upstream_other_npdes_count',
  'stationcategory',
  'month'
)
```

Note that huc10id is not included in the random effects in the following models because there is a high degree of confounding between the huc10's and stations.

### Nitrite/nitrate concentration

Best predictor: huc12_nearest_upstream_cafo_dist_scaled

```{r}
nitlog_results <- cafo_model_selection(huc12_df,
                                       response = 'nitlog',
                                       all_cafo_predictors,
                                       confounders)
nitlog_results
```

Best formula: nitlog \~ huc12_nearest_upstream_cafo_dist_scaled + stationcategory + month + (1 \| huc8id/station_id)

```{r}
summary(nitlog_results$best_model)
```

### Stepwise selection for fecal coliform count

Best predictor = huc12_nearest_upstream_cafo_dist_scaled

```{r}
feclog_results <- cafo_model_selection(huc12_df,
                                       response = 'feclog',
                                       all_cafo_predictors,
                                       confounders)
feclog_results
```

Best model

formula: feclog \~ huc12_nearest_upstream_cafo_dist_scaled + huc12_upstream_other_npdes_count +\
stationcategory + month + (1 \| huc8id/station_id)

```{r}
summary(feclog_results$best_model)
```

### Stepwise selection for phosphorus

In this model, including huc10id did not result in a singular fit warning, but it did not improve the model, so it was excluded for consistency.

best predictor = huc12_nearest_upstream_cafo_dist_scaled

```{r}
phoslog_results <- cafo_model_selection(huc12_df,
                                       response = 'phoslog',
                                       all_cafo_predictors,
                                       confounders)
phoslog_results
```

Best model

Formula: phoslog \~ huc12_nearest_upstream_cafo_dist_scaled + huc12_upstream_other_npdes_count +\
stationcategory + month + (1 \| station_id:huc10id:huc8id) + (1 \| huc8id)

```{r}
summary(phoslog_results$best_model)
```

huc12_nearest_upstream_cafo_dist_scaled was the best predictor for all of the models tested above.

## Stepwise Selection: Separate CAFO Metrics Models

Prepare predictors and confounding variables

Can't include distance variables for other and swine CAFO's because NA values will be dropped and only one unique station_id will be left

```{r}
huc12_poultry_df <- df |>
  drop_na(huc12_nearest_upstream_poultry_c)
  
huc12_poultry_df <- huc12_poultry_df |>
  mutate(
    huc12_upstream_p_cafo_density_scaled = huc12_upstream_poultry_cafo_dens / 0.1,
    huc12_upstream_p_cafo_count_scaled = huc12_upstream_poultry_cafo_coun / 10,
    huc12_weighted_p_cafo_load_scaled = huc12_weighted_poultry_cafo_load / 10,
    huc12_nearest_p_upstream_cafo_dist_scaled = huc12_nearest_upstream_poultry_c / 1000,
    
    # other cafo confounding vars
    huc12_upstream_other_cafo_density_scaled = huc12_upstream_other_cafo_densit / 0.1,
    huc12_upstream_other_cafo_count_scaled = huc12_upstream_other_cafo_count / 10,
    
    # swine cafo confounding vars
    huc12_upstream_s_cafo_density_scaled = huc12_upstream_swine_cafo_densit / 0.1,
    huc12_upstream_s_cafo_count_scaled = huc12_upstream_swine_cafo_count / 10
    
  )
```

```{r}
# scaled cafo predictors to test one at a time
p_cafo_predictors <- c(
  'huc12_upstream_p_cafo_density_scaled',
  'huc12_upstream_p_cafo_count_scaled',
  'huc12_weighted_p_cafo_load_scaled',
  'huc12_nearest_p_upstream_cafo_dist_scaled'
)

# confounders for backward elimination
p_confounders <- c(
  'huc12_upstream_other_cafo_density_scaled',
  'huc12_upstream_other_cafo_count_scaled',
  'huc12_upstream_s_cafo_density_scaled',
  'huc12_upstream_s_cafo_count_scaled',
  'huc12_pop_density',
  'huc12_pct_cropland',
  'huc12_upstream_wastewater_count',
  'huc12_upstream_landfill_count',
  'huc12_upstream_other_npdes_count',
  'stationcategory',
  'month'
)
```

### Nitrite/nitrate Concentration

-   Need to figure out where the 'simpleWarning: Some predictor variables are on very different scales: consider rescaling.' warnings are coming from and scale those variables appropriately

```{r}
nitlog_p_results <- cafo_model_selection(huc12_poultry_df,
                                       response = 'nitlog',
                                       predictors = p_cafo_predictors,
                                       p_confounders)
nitlog_p_results
```

Best formula: nitlog \~ huc12_nearest_p_upstream_cafo_dist_scaled + month + (1 \| station_id:huc8id)

```{r}
summary(nitlog_p_results$best_model)
```

### Fecal Coliform Counts

```{r}
feclog_p_results <- cafo_model_selection(huc12_poultry_df,
                                       response = 'feclog',
                                       predictors = p_cafo_predictors,
                                       p_confounders)
feclog_p_results
```

Best formula: feclog \~ huc12_upstream_p_cafo_density_scaled + huc12_pct_cropland +\
huc12_upstream_landfill_count + huc12_upstream_other_npdes_count + month + (1 \| huc8id/station_id)

```{r}
summary(feclog_p_results$best_model)
```

### Phosphorous Concentration

```{r}
phoslog_p_results <- cafo_model_selection(huc12_poultry_df,
                                          response = 'phoslog',
                                          predictors = p_cafo_predictors,
                                          p_confounders)
phoslog_p_results
```

Best formula: phoslog \~ huc12_upstream_p_cafo_density_scaled + huc12_upstream_s_cafo_density_scaled +\
huc12_upstream_s_cafo_count_scaled + huc12_pct_cropland + stationcategory + month + (1 \| station_id:huc8id)

```{r}
summary(phoslog_p_results$best_model)
```

## To-do:

-   Test models with separate metrics for poultry, swine, and other CAFOs in same model
-   Fit selected models and evaluate their performance
-   Sensitivity analysis
